Distributional Trees for Circular Data

### 2019-01-31 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revisions
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2415 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ccd205def5da9487d04e8ac86b6c3618a1186e61

\section{Motivation}
Circular response variables occur in a variety of applications and subject areas. 
For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
animal orientation or gene-structure analysis are often subject of examination in biology, 
and wind direction is one of the most important weather variables in meteorology.
Representing such a two-dimensional direction by an angle in a unity circle requires 
the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
rotation. However, these settings can vary but should not influence statistical 
inference from a fitted model.
For that reason circular data demand for purpose-built statistical methods. 
One common way of modelling circular data is to apply the von Mises distribution 
which is often regarded as the circular analog of the normal distribution. 
Estimating its two distribution parameters by generalized additive models 
using a set of regressor variables is one possible approach that can be successful 
in various settings.
As an alternative we propose the application of distributional trees 
which are capable of detecting also non-additive effects and provide easily interpretable 
models allowing for a wide range of inference procedures.
Moreover, covariates and their possible interactions are selected automatically
such that no expert knowledge is required to specify the model in advance.

@@

\subsection{Von Mises Distribution}
A commonly chosen distribution to model circular data is the von Mises distribution
which is also known as the circular normal distribution 
(Jammalamadaka and Sengupta, 2001; Mardia an dJupp, 2009). 
Its density function is given by
\begin{equation}
  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
\end{equation}
with a location parameter $\mu$, a dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
modified Bessel function of the first kind and order $0$.
For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
to the variance $\sigma^2$ of the normal distribution.

\subsection{Distributional Trees}
Distributional trees (Schlosser et. al, 2018) embed distributional modeling such as
provided by generalized additive models for location, scale, and shape 
(GAMLSS, Rigby and Stasinopoulos, 2005) into the framework of regression trees (Breiman et. al, 1984).
In particular one of the unbiased recursive partitioning algorithms MOB (Zeileis et. al, 2008) 
and CTree (Hothorn et. al, 2006) is applied and a distributional model is fitted in each node 
by estimating the parameters of a specified distribution family via maximum likelihood.
For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
tests applied on model scores and each possible split variable. After choosing the split variable 
showing the highest association to the goodness of fit of the model the corresponding split point 
which optimizes an objective function is selected in a separate step.

In the here discussed application the MOB algorithm is applied and the prespecified 
distribution family is the von Mises distribution. 
The corresponding distribution parameter vector $\theta = (\mu,\kappa)$ is estimated 
by maximizing $l(y, \theta) = log(f_{vM}(y;\theta))$ and model scores are calculated 
by $s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. In that way
a value indicating the goodness of fit is provided for each parameter and observation
such that changes can be detected not only in the location parameter $\mu$ but also in the 
dispersion parameter $\kappa$.

@@

%\begin{figure}[!ht]\centering
%\includegraphics[width = 1.\textwidth]{circtree_plot_small.pdf}
%\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises distribution.}\label{fig:tree}
%\end{figure}

@@

\section{Outlook}

While the application of the von Mises distribution allows for modelling a circular response variable,
additional adaptions have to be made when including circular regressor variables in a distributional tree
model. In particular, if the selected split variable is a circular variable this demands for finding
not only one split point but a ``split interval''. This consideration is of interest for many applications,
for example when modelling wind directions it might be desirable to include day time or earlier 
observations of wind directions as regressor variables.

@@

\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
     {\it working paper}.

>>>>

\section{Motivation}
Circular response variables occur in a variety of applications and subject areas. 
For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
animal orientation or gene-structure analysis are often subject of examination in biology, 
and wind direction is one of the most important weather variables in meteorology.
Representing such a two-dimensional direction by an angle in a unity circle requires 
the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
rotation. However, these settings can vary but should not influence statistical 
inference from a fitted model.
For that reason circular data demand for purpose-built statistical methods. 
A commonly chosen distribution for modeling circular data is the von Mises distribution.
Estimating its two distribution parameters by generalized additive models using a set of 
regressor variables is one possible approach that can be successful in various settings.
As an alternative we propose the application of distributional trees 
which are capable of detecting also non-additive effects and provide easily interpretable 
models allowing for a wide range of inference procedures.
Moreover, covariates and their possible interactions are selected automatically
such that no expert knowledge is required to specify the model in advance.

@@

%\subsection{Von Mises Distribution}
%The von Mises distribution is also known as the circular normal distribution 
%and its density function is given by
%\begin{equation}
%  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
%\end{equation}
%with a location parameter $\mu$, a dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
%modified Bessel function of the first kind and order $0$.
%For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
%a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
%to the variance $\sigma^2$ of the normal distribution. 
%See e.g., Jammalamadaka and Sengupta (2001) or Mardia and Jupp (2009) for a more detailed overview.

%\subsection{Distributional Trees}
%Distributional trees~(Schlosser et. al, 2018) embed distributional modeling into the framework 
%of regression trees.
Distributional trees~(Schlosser et. al, 2018) embed distributional modeling such as
provided by generalized additive models for location, scale, and shape~(GAMLSS, 
Rigby and Stasinopoulos, 2005) into the framework of regression trees~(Breiman et. al, 1984).
In particular one of the unbiased recursive partitioning algorithms MOB~(Zeileis et. al, 2008) 
and CTree~(Hothorn et. al, 2006) is applied and a distributional model is fitted in each node 
by estimating the parameters of a specified distribution family via maximum likelihood.
For each split the split variable is selected based on $p$-values from permutation or M-fluctuation 
tests evaluating associations between model scores and each possible split variable. After choosing 
the split variable showing the highest association to the goodness of fit of the model the corresponding split point is selected in a separate step by optimizing an objective function.

In the here discussed application the MOB algorithm is applied and the prespecified 
distribution family is the von Mises distribution which is also known as the circular 
normal distribution. Its density function is given by
\begin{equation}
  f_{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{equ:vm}
\end{equation}
with a location parameter $\mu$, a dispersion parameter $\kappa$, and $I_0(\kappa)$ being the 
modified Bessel function of the first kind and order $0$.
For $\kappa=0$ a uniform distribution is obtained. Otherwise, a large value of $\kappa$ indicates
a high concentration of the distribution around $\mu$. Hence, $\frac{1}{\kappa}$ is analogous
to the variance $\sigma^2$ of the normal distribution. 
(See e.g., Jammalamadaka and Sengupta (2001) or Mardia and Jupp (2009) for a more detailed overview.)
The distribution parameter vector $\theta = (\mu,\kappa)$ is estimated 
by maximizing $l(y, \mu, \kappa) = log(f_{vM}(y;\mu, \kappa))$ and model scores are calculated 
by 
$s(y,\hat{\mu}, \hat{\kappa}) = \left(\frac{\partial l(y, \hat{\mu}, \hat{\kappa}}{\partial \mu},
\frac{\partial l(y, \hat{\mu}, \hat{\kappa}}{\partial \kappa}\right)$. 
%$s(y,\hat{\theta}) = \frac{\partial l(y, \hat{\theta})}{\partial \theta}$. 
In that way
a value indicating the goodness of fit is provided for each parameter and observation
such that changes can be detected not only in the location parameter $\mu$ but also in the 
dispersion parameter $\kappa$.

@@

\begin{figure}[!ht]\centering
\includegraphics[width = 1.\textwidth]{circtree_plot_small.pdf}
\caption{\label{fig:circtree} Fitted distribution tree for a circular response employing the von Mises distribution.}\label{fig:tree}
\end{figure}

@@

%\bigskip
\section{Outlook}

While the application of the von Mises distribution allows for modeling a circular response variable,
additional adaptions have to be made when including circular regressor variables in a distributional tree
model. In particular, if the selected split variable is a circular variable this demands for finding
not only one split point but a ``split interval''. This consideration is of interest for many applications,
for example when modeling wind directions it might be desirable to include day time or earlier 
observations of wind directions as regressor variables.

@@

\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
     {\it arXiv.org E-Print Archive}, arXiv 1804.02921, \url{http://arxiv.org/abs/1804.02921}
